Advanced Data Structures - Succinct Bit Vector

### 2024-07-03 article/article.tex
add algorithm implementation

% \pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
\nolinenumbers

@@

\section{Algorithm}
In the following section we will present an interleaved datastructure
that implements \code{rank}, \code{select} and \code{access} operations.
The interleaved data structure aims to use the CPU cache as efficiently as possible, where both
vector information and metadata should fit in the same cache line of 512 bit.
Our bit vector implementation consists of \code{uint64} blocks, of which we can fit 8 inside one cache line
$L_i = (\prevsum, \vec{v}_0, \dots, \vec{v}_6)$.
To speed up the computation of rank and select we need metadata, which we store in the first
64 bits of the line.
This data structure is depicted in Figure~\ref{fig:interleaved}. 
This leaves us with a memory overhead of $12.5\%$. The used metadata $\prevsum$ is an incremental sum of previous bits
set to one. For an efficient build time of the data structure we make use of the pop count $c(\vec{v})$
instruction, which counts the number of 1 bits in a subvector $\vec{v}$, in our case an \code{uint64} number.

@@

\section{Evaluation}
In this chapter we will evaluate our predicted runtime against measurements.
For the pre-computation we calculate an incremental sum $\prevsum$ every 7 blocks of 64 bits.
This should scale linearly with the vector size $|\vec{V}|=n$.
As for \code{rank} we can calculate the position of the vector line in constant time.
For \code{select} we need to first run a binary search in $\mathcal{O}(\log{n})$ to find the vector line.
After that we compute a linear search in $\mathcal{O}(7)$ as there is a maximum of 7 blocks we need to search.
The runtime should be dominated by the binary search.
To validate these predictions we conducted an experiment were a bitvector $\vec{V}$ with increasing
size was randomly generated as well as a fixed amount of $100 000$ commands.
We then measured the runtime for the pre-computation as well as the command execution.
In Figure~\ref{fig:performance_eval} the recorded data is visualized.
\begin{figure}[htbp]
    \centering
    \input{images/benchmark.tex}
    \caption{%
        Performance Evaluation of an increasing vector size and a constant 100 000 commands.
        The benchmark was run on both:
        a 12th Gen Intel® Core™ i5-1240P × 16 with 32 GiB RAM, 1 TB SSD running Fedora Linux 40
        and on a dedicated vCPU Server (Hetzner CCX33) with 
        8 Cores, 32 GiB RAM and 240 GiB SSD running on AMD Milan EPYC™ 7003 or AMD Genoa EPYC™ 9654
        running Ubuntu 24.04.
    }
    \label{fig:performance_eval}
\end{figure}
Analyzing our data we can confirm the predicted runtime, where the pre-computation scales linearly
with the vector length, and the command execution time scaling logarithmic.
We also notice that the pre-computation on the AMD setup being faster while the command execution being
slower in comparison to the Intel setup.

>>>>

% \pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
% \nolinenumbers

@@

\section{Algorithm}
In the following section we will present a datastructure
that implements \code{rank}, \code{select} and \code{access} operations.
The data structure aims to use the CPU cache as efficiently as possible, by interleaving both
vector information and metadata to fit in the same cache line of 512 bit.
Our bit vector implementation consists of \code{uint64} blocks, of which we can fit 8 inside one cache line
$L_i = (\prevsum, \vec{v}_0, \dots, \vec{v}_6)$.
To speed up the computation of rank and select we need metadata, which we store in the first
64 bits of the line.
This data structure is depicted in Figure~\ref{fig:interleaved}. 
This leaves us with a memory overhead of $12.5\%$. The used metadata $\prevsum$ is an incremental sum of previous bits
set to one. For an efficient build time of the data structure we make use of the pop count $c(\vec{v})$
instruction, which counts the number of 1 bits in a subvector $\vec{v}$, in our case an \code{uint64} number.

@@

\section{Evaluation}
In this chapter we will evaluate our predicted runtime against measurements.
For the pre-computation we calculate an incremental sum $\prevsum$ every 7 blocks of 64 bits.
This should scale linearly with the vector size $|\vec{V}|=n$.
As for \code{rank} we can calculate the position of the vector line in constant time.
For \code{select} we need to first run a binary search in $\mathcal{O}(\log{n})$ to find the vector line.
After that we compute a linear search in $\mathcal{O}(7)$ as there is a maximum of 7 blocks we need to search.
The runtime should be dominated by the binary search.
To validate these predictions we conducted an experiment were a bitvector $\vec{V}$ with increasing
size was randomly generated as well as a fixed amount of $100 000$ commands.
We then measured the runtime for the pre-computation as well as the command execution.
In Figure~\ref{fig:performance_eval} the recorded data is visualized.
\begin{figure}[t]
    \centering
    \input{images/benchmark.tex}
    \caption{%
        Performance Evaluation of an increasing vector size and a constant 100 000 commands.
        The benchmark was run on both:
        a 12th Gen Intel® Core™ i5-1240P × 16 with 32 GiB RAM, 1 TB SSD running Fedora Linux 40
        and on a dedicated vCPU Server (Hetzner CCX33) with 
        8 Cores, 32 GiB RAM and 240 GiB SSD running on AMD Milan EPYC™ 7003 running Ubuntu 24.04.
    }
    \label{fig:performance_eval}
\end{figure}
Analyzing our data we can confirm the predicted runtime, where the pre-computation scales linearly
with the vector length, and the command execution time scaling logarithmic.
We also notice that the pre-computation on the AMD setup being faster while the command execution being
slower in comparison to the Intel setup.
